---
title: "FA3_Cuerdo, Naomi Hannah A."
author: "Cuerdo, Naomi Hannah A."
date: "2025-02-22"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Major League Basketball

```{r libraries, include= FALSE}
library(tidyverse)
library(ggrepel)
library(ggplot2)
library(kableExtra)
library(patchwork)
```

## Import the Data

```{r loading data, include=FALSE}
load("C:/Users/naomi/Downloads/ml_pay.rdata")
ls()
```

```{r}
mlb_raw <- as_tibble(ml_pay)
print(mlb_raw)
```
The table contains 30 rows and 54 columns which matches the data description.


## Tidying the Data
```{r tidy}
payroll_long <-mlb_raw %>%
    pivot_longer(
    cols = matches("^p\\d{4}$"), 
    names_to = "Year",
    names_prefix="p",
    values_to = "Payroll_p_year"
  )
  
wins_long <- mlb_raw %>%
    pivot_longer(
    cols = matches("^p\\d{4}$"), 
    names_to = "Year",
    names_prefix="X",
    values_to = "num_wins"
  )

wins_pct_long <- mlb_raw %>%
    pivot_longer(
    cols = matches("^X\\d{4}\\.pct$"), 
    names_to = "Year",
    names_prefix="X",
    values_to = "pct_wins"
  ) %>% 
  mutate(Year = str_remove(Year, "\\.pct$"))

df_list <-list(payroll_long, wins_long, wins_pct_long)

tidy_data <- reduce(df_list, left_join, by = c("Team.name.2014", "Year"))

tidy_data <- tidy_data %>%
  select(-matches("^p\\d{4}$"), -matches("^X\\d{4}$"), -matches("^X\\d{4}\\.pct$"))

tidy_data <- tidy_data %>%
  select(-matches("\\.x$")) %>%
  select(-matches("\\.y$"))

tidy_data <- tidy_data %>%
  mutate(Payroll_p_year = as.numeric(Payroll_p_year))


tidy_data <- tidy_data %>%
  rename(team = Team.name.2014)

tidy_data


```

Now that the data has been tidy, we need to create two more tables: **mlb_aggregate and mlb_yearly** using **tidy_data** 

```{r mlb_aggregate}

mlb_aggregate <-mlb_raw %>%
  select(Team.name.2014, payroll, avgwin) %>%
  rename(team = Team.name.2014)

mlb_aggregate
```

```{r mlb_yearly}

mlb_yearly <-tidy_data %>%
  select(-payroll, -avgwin)

mlb_yearly

```

**mlb_aggregate** has 30 rows, while **mlb_yearly** has 510 rows. **mlb_yearly** has more rows than **mlb_aggregate** since **mlb_yearly** records the wins of teams year by year, percentage, and its payroll per year. The **mlb_aggregate** only shows the total payroll and average wins. 

## Quality Control
```{r qual control compu}
mlb_aggregate_computed <- mlb_yearly %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate_computed = sum(Payroll_p_year, na.rm = TRUE),
    pct_wins_aggregate_computed = mean(pct_wins, na.rm = TRUE)
  )

mlb_aggregate_computed

```


```{r joing the tibbles}
mlb_aggregate_joined <- mlb_aggregate %>%
  left_join(mlb_aggregate_computed, by = "team")

mlb_aggregate_joined


```


Visualizing **mlb_aggregate_joined** with scatter plots, comparing **payroll_aggregate_computed** vs **payroll_aggregate**, and **pct_wins_aggregate_computed vs pct_wins_aggregate**

```{r scatter plot}

payroll_scatter <- ggplot(mlb_aggregate_joined, aes(x=payroll_aggregate_computed, y = payroll, label = team)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Aggregated vs. Provided Payroll", x = "Aggregated Payroll", y = "Provided Payroll")

win_pct_scatter <- ggplot(mlb_aggregate_joined, aes(x=pct_wins_aggregate_computed, y = avgwin, label = team)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Aggregated vs. Provided Win %", x = "Aggregated Win %", y = "Provided % Win")

payroll_scatter + win_pct_scatter

```

Comparing the two graphs above, we can conclude that the computed payroll and win percentage have no significant difference, indicating that no errors are present in the computed payroll and win percentage.

## Visualization

### Payroll across years

```{r payroll across years}
payroll_year <- ggplot(tidy_data, aes(x = Year, y = Payroll_p_year, group = team, color = team)) +
  geom_line() +
  geom_hline(aes(yintercept = mean(Payroll_p_year)), linetype = "dashed", color = "purple") +
  facet_wrap(~ team, scales = "free_y") +
  labs(title = "Payroll Across Years", x = "Year", y = "Payroll") +
  theme_minimal()

payroll_year
```

Identifying the the three teams with the greatest payroll_aggregated computed:

```{r greatest payroll}
top_payroll_teams <- mlb_aggregate_computed %>%
  arrange(desc(payroll_aggregate_computed)) %>%
  slice_head(n = 3)

top_payroll_teams

```
The Top Three teams are: New York Yankees (1st), Boston Red Sox (2nd), and Los Angeles Dodgers (3rd).


Identifying the three teams with the greatest percentage increase in payroll from 1998 to 2014:

```{r percentage}

payroll_growth <- mlb_yearly %>%
  filter(Year  %in% c(1998, 2014)) %>%
  select(team, Year, Payroll_p_year) %>%
  pivot_wider(names_from = Year, values_from = Payroll_p_year, names_prefix = "Payroll_") %>%
  mutate(pct_increase = (Payroll_2014 - Payroll_1998) / Payroll_1998 * 100) %>%
  arrange(desc(pct_increase)) %>%
  slice_head(n = 3)

payroll_growth


```

The Top Three teams with the greatest increase in payroll from 1998 to `2014 are: Washington Nationals, Detroit Tigers, and Philadelphia Phillies.


The top 3 payroll teams have been having consistent high payroll per year, which can be seen in the plot. The top 3 also are lying above the average of the payrolls

The 3 teams for percentage increase have a more steeper increase. 

### Win Percentage Across Years

```{r win percentage }
pct_wins_plot <- ggplot(tidy_data, aes(x = Year, y = pct_wins, group = team, color = team)) +
  geom_line() +
  geom_hline(aes(yintercept = mean(pct_wins)), linetype = "dashed", color = "purple") +
  facet_wrap(~ team, scales = "free_y") +
  labs(title = "Win Percentage Across Years", x = "Year", y = "pct_wins") +
  theme_minimal()

pct_wins_plot

```
Identifying the first three teams with the greatest **pct_wins_aggregate_computed** 

```{r top three pct wins}
top_pct_teams <- mlb_aggregate_computed %>%
  arrange(desc(pct_wins_aggregate_computed)) %>%
  slice_head(n = 3)

top_pct_teams

```

The top three teams with the highest payroll are: New York Yankees, Atlanta Braves, and Boston Red Sox.


Identifying the three teams with the most erratic percentage wins across years measured by standard deviation

```{r pct wins sd}

sd_teams <- mlb_yearly %>%
  group_by(team) %>%
  summarize(pct_wins_sd = sd(pct_wins, na.rm = TRUE)) %>%
  arrange(desc(pct_wins_sd)) %>%
  slice_head(n = 3)

sd_teams


```

The top three teams with the erratic percentage in standard deviation are: Houston Astros, Detroit Tigers, and Seattle Mariners.

The results from the **dpylr** amd from the plot coincide in the top 3 percentage. As an example, the New York Yankees had the highest payroll account to 70%, but it dropped by 59% as years go by.

Similarly, in the top 3 teams with the highest standard deviation, Houston Astros has the highest sd, which can be seen in the plot that goes down as time goes by. 

### Win Percentage versus Payroll

```{r}
win_pct_payroll <-  ggplot(mlb_aggregate_joined, aes(x = payroll_aggregate_computed, y = pct_wins_aggregate_computed, label = team)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_text_repel(max.overlaps = Inf) +
  theme_minimal() +
  labs(title = "Payroll vs. Win Percentage", x = "Aggregated Payroll", y= "Aggregated Win Percentage")

win_pct_payroll
```

From the graph above, it seems that there is a positive relationship between the aggregated payroll and the aggregated win percentage of the teams. As  the aggregated payroll increases, the aggregated win percentage also increases.

## Team Efficiency


```{r team efficiency}

mlb_aggregate_joined <- mlb_aggregate_joined %>%
  mutate(Efficiency = pct_wins_aggregate_computed / payroll_aggregate_computed )

team_efficiency <- ggplot(mlb_aggregate_joined, aes(x = reorder(team, -Efficiency), y = Efficiency, fill = team)) +
  geom_col() +
  labs(title = "Team Efficiency", x = "Team", y = "Efficiency (Win % per Payroll)") +
  theme(axis.text.x = element_text(angle= 90, hjust = 1))

team_efficiency


```

Identifying the first three teams with the greatest efficiency:

```{r great efficiency}

top_teams <- mlb_aggregate_joined %>%
  arrange(desc(Efficiency)) %>%
  slice_head(n = 3)

top_teams

```

The top three efficient teams: Miami Marlins, Oakland Athletics, and Tampa Bay Rays. These three teams are efficient in terms of their winrate based on payroll. 
