---
title: "FA1"
author: "Cuerdo, Naomi Hannah A."
date: "2025-02-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lecture 2

```{r libraries, include = FALSE}
library(tidyverse)
library(ggcorrplot)
```

## Case Study: Diamonds
### Loading the Library
```{r load libraries}
diamonds
```

## 3 Variations

### 3.1 Discrete Variable

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = (aes(x = cut)))
```

Figure 1 above shows a bar graph of the distribution of different diamond cuts, categorized as **Fair, Good, Very Good, Premium, and Ideal**. The y-axis represents the count of diamonds for each cut, providing insight into the most and least common types of diamond cuts. 

From Figure 1,**Ideal** is the most common cut, with a frequency of 20,000 diamonds, followed by **Premium** and **Very Good**. This suggests that **ideal-cut diamonds** are the most preferred, possibly due to their brilliance and market demand. **Good** cut is moderately frequent, with a frequency of 5,000 diamonds. 

On the other hand, The graph also shows that fair cut is the least common cut, suggesting that diamonds are either less desirable or less frequently produced. This trend suggests a strong market preference for higher-quality cuts, likely due to their superior brilliance and appeal. 

### 3.2 Continuous variables

```{r cont}
ggplot(data = diamonds) +
geom_histogram(mapping = aes(x = carat))

```
Figure 2 about shows a histogram of the frequency of carat distributions. From figure 2, it shows that most of the values of carat are below 3, so it is best to zoom in to the portion of the plot. 

```{r zoom }
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat)) +
  coord_cartesian(xlim = c(0,3))

```
Figure 3 seems to have an error with the bin width. Decreasing the bin width of the graph for better visualization:

```{r graph 3}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.01)
coord_cartesian(xlim = c(0,3))

```

Figure 4 shows the distribution of carat values. The x-axis represents the **carat values**, while the y-axis represents the count of diamonds for each bin, with a bin width of 0.01. 

Most of the diamonds have small carat values below 1. The distribution is **right-skewed**. There are noticeable spikers at common carat sizes (e.g., **0.25, 0.5, 0.75, 1.0 carats **), suggesting certain standardized weights are more frequently produced or preferred by consumers. As carat increases beyond **2 carats**, the count drops significantly, indicating that **large diamonds are much rarer.**


## 4 Covariation
### 4.1 Discrete versus discrete.
```{r discrete}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = clarity))

```

The bubble chart above (Figure 5) between **diamond cut** and **clarity**, where the size of each bubble represents the count of diamonds in each category. 

From the chart, **Ideal cut dominates**, wherein most diamonds, across all clarity levels have an **ideal cut**, suggesting high consumer preference. **Premium** and *Very Good* cuts are also common, where they have substantial representation across different clarity grades. Fewer diamonds with **Fair** and **Good ** cuts appear less frequently.

Overall, **clarity distribution is fairly even**, suggesting that no single clarity grade overwhelmingly dominates any cut category, though **VS2** and **SI1** seem more common overall. 


### 4.2 Continuous versus continuous
#### 4.2.1 Many to one
```{r scatter plot}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y))


```

omitting the outliers, we get:
```{r scat no outliers}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10))

```

From the graph, it seems that most of the points lie on the line y = x. To confirm this, we shall add a line:

```{r}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_hline(yintercept = 6, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = 6, color = "green", linetype = "dashed") + 
  coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10)) 
```

Figure 6 shows a scatter plot of the relationship between the x (length)and y (width) dimensions of diamonds in the data set. The red diagonal line helps assess how closely x and y values match. The blue and green dashed lines serve as thresholds to categorize diamonds based on their dimensions. 

The points near the red line indicate that diamonds have nearly equal length and width. The diamonds that width greater than 6 units are above the red line, while those  smaller are below it. The green line shows the diamonds with greater length than 6 units are to the right, while the smaller ones are to the left. 

Overall, the graph offers a clear picture of the relationships between length and width of the diamonds, which gives us a result of equal sizes in both dimensions. 

#### 4.2.2 One to one
We now then proceed to visualize the average diamond price for each value of the carat.

```{r avg price}
avg_price_by_carat <- diamonds %>%
  mutate(carat = round(carat, 1)) %>%
  group_by(carat) %>%
  summarise(avg_price = mean(price))
avg_price_by_carat

```

Plotting the avg_price versus carat using a line plot:

```{r line plot} 
ggplot(data = avg_price_by_carat) +
  geom_line(mapping = aes(x = carat, y = avg_price))

```
Figure 7 shows a line graph how diamond prices increase with carat size. From the plot, it shows that the price peaks at carat size 5, with a price range of around 20,000. However, The price dropped at sizes drop at around size 3 and 4, which indicates that the line is not entirely smooth as there are fluctuations with the price increase. 

The graph makes sense since larger diamonds tend to be rarer and more desirable. The fluctuations may be due to **data variability, outliers, or market demand for certain carat sizes.**

### 4.3 Continuous versus discrete

```{r cvsd}
ggplot(data = diamonds) + 
  geom_boxplot(mapping = aes(x = cut, y = price))
```

### 4.4 Correlation matrices

Computing for the correlation matrix:

```{r corre matrices}
corrmat <- cor(diamonds[, sapply(diamonds, is.numeric)], use = "complete.obs")
```

```{r corre matrices 1}
corrmat
```

```{r corre matrices 2}
ggcorrplot(corrmat, lab=TRUE, hc.order = TRUE)

```
Figure 8 above reveals strong positive relationships between carat, price, and the dimensions (x, y, z), with correlations **0.95-0.98**. 

This plot indicates that larger diamonds tend to be more expensive in carat weight and physical dimensions. The dimensions x, y, and z are also highly correlated with each other, which serves as a reference of carat and its price.

In contrasts, **table and depth** show much weaker relationships with price, with having a **low correlation of 0.13** and **depth showing almost no correlation (-0.01)**. Additionally, there is a **moderate negative correlation (-0.3)** between **table and depth**, suggesting an inverse relationships in how diamonds are cut. 

Overall, **carat and dimensions** are the primary source of its price, while **depth and table** have minimal influence. 


## 5 Additional visualization tools

### 5.1 Axis transformations

```{r axis trans 1}
ggplot(data = diamonds) +
geom_histogram(aes(x = carat), binwidth = 0.01) +
scale_x_log10()

```

Figure 9 shows a histogram of carats size in a logarithmic scale for the x axis. The carat distribution is right-skewed, meaning most diamonds have smaller carat values, with fewer large-carat diamonds.

The plot uses a log10 scaling that helps spread out smaller values, making the patterns and density more interpretable.  

Furthermore, also shows that carat size 1.0 is the most popular, followed by 0.5 and 1.5. 

Plotting the variables on the y-axis:
```{r y axis}
ggplot(data = diamonds) +
geom_point(aes(x = carat, y = price)) +
scale_x_log10() +
scale_y_log10()
```

Figure 10 shows a scatter plot between variables carat (x-axis) and price (y-axis). This plot further solidifies that carats with much higher prices have minimal occurrences in the graph, which indicates that carats with larger sizes are rare to find.

### 5.2 Multiple geoms in the same plot

```{r}
ggplot(data = avg_price_by_carat, mapping = aes(x = carat, y = avg_price)) +
geom_line() +
geom_point()
```

Figure 11 shows a line graph how diamond prices increase with carat size. From the plot, it shows that the price peaks at carat size 5, with a price range of around 20,000. However, The price dropped at sizes drop at around size 3 and 4 to around 10,000 and 15, 0000, which indicates that the line is not entirely smooth as there are fluctuations with the price increase. 

### 5.3 Multiple aesthetics in the same geom
adding a color for the scatter plot:

```{r aesthetic}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut))

```

Figure 12 shows a plot between the relationship between the carat and price, with data points categorized by cut quality. 

From the plot, there is a **clear positive correlation**, where price generally increases with carat size. Higher-quality cuts, such as **"Ideal" and "Premium,"** tend to be priced higher, even at lower carat sizes, indicating that cut significantly impacts value. The distribution of points shows that **smaller diamonds are more common, while larger diamonds are rarer and more expensive.**

Additionally, some outliers suggest that factors beyond carat and cut, such as clarity and color, influence pricing. Overall, the graph highlights the complex pricing structure of diamonds, where both size and quality contribute to their market value.

### 5.4 Faceting
```{r faceting 1}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) +
  facet_wrap(~cut)

```

```{r faceting 2}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) +
  facet_grid(. ~ cut)

```
Figures 12 and 13 show scatter plots between carat and its price per cut. The plot shows that fair cuts are much frequent in the lower price range, while higher prices are much frequent in very good, premium, and ideal. The ideal cut, however, tends to be more frequent in all price ranges, indicating that the ideal cut is the most common in diamonds. 

```{r faceting 3}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) +
  facet_grid(clarity ~ cut)

```

Figure 15 shows a faceted scatter plot of how diamond price varies with carat size, categorized by **cut quality (columns)** and **clarity (rows).** Price generally increases with carat, but higher cut quality **(Ideal, Premium)** and **better clarity (IF, VVS1, VVS2)** result in significantly higher prices, while **lower clarity and cut grades (e.g., I1, Fair)** tend to have lower prices even for larger carats.

The plot highlights how carat, cut, and clarity collectively influence diamond pricing, with premium cuts and higher clarity diamonds commanding the highest values.

## 7 Exercises
### 7.1 Relating the carat and cut of a diamond
```{r carat and cut}
ggplot(diamonds, aes(x = cut, y = carat, fill = cut)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Distribution of Carat Size by Cut Quality",
       x = "Cut Quality",
       y = "Carat Size") +
  scale_fill_brewer(palette = "Pastel1")

```

The first scatter plot shows that while price generally increases with carat size, higher-quality cuts (Ideal, Premium) are more common in smaller carat ranges, whereas lower-quality cuts (Fair, Good) extend to larger carat sizes. This suggests an inverse relationship between carat and cut quality, where larger diamonds often have lower cut grades.

A box plot of carat size by cut category would further confirm this trend, showing that diamonds with better cuts tend to be smaller. This explains the paradox in the first plot, where some smaller diamonds are priced higher than larger ones. 
