---
title: "FA1_Lecture 3-Cuerdo, Naomi Hannah A."
author: "Cuerdo, Naomi Hannah A."
date: "2025-02-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyverse)
library(dplyr)
diamonds
```

# Lecture 3

## 2 Isolating Data
### 2.1 filter()

```{r filter1}
filter(diamonds, price >= 10000)
```

filtering diamonds with price at least $10,000 and clarity VVS1 or IF:
```{r filter2}
filter(diamonds, price >10000 & clarity %in% c("VVS1", "IF"))

```
### 2.2 select()
```{r select}
select(diamonds, carat, cut, color, clarity)

```

### 2.3 arrange()
sorting diamonds by carat in ascending order:

```{r arrange ascending}

arrange(diamonds,carat)

```

arranging diamonds by carat in descending order:
```{r arrange descending}
arrange(diamonds, desc(carat))

```

Arranging diamonds in decreasing order of their length:
```{r arrange length}
arrange(diamonds, desc(x))
```
From the table above, the longest diamond is 5.01 fair cut carat, with a length of 10.74.

## 3 Deriving information
### 3.1 mutate()

adding a column that is the price per carat of each diamonds:

```{r mutate}
mutate(diamonds, price_per_carat = price/carat)

```

adding a column that indicates whether a diamonds price per carat is at least $10,000:
```{r carat }
mutate(diamonds, fancy_diamond = price/carat > 10000)

```

Checking the quality of the diamond based on its size:
```{r}
mutate(diamonds, 
       value =
         case_when(
           carat > 2 & price < 5000 ~ "good", # if carat > 2 and price < 5000, then good
           carat > 1 & price < 2500 ~ "ok", # if carat > 1 and price < 2500, then ok
           TRUE ~ "bad" # otherwise, bad
           )
       )
```

### 3.2 summarise()
finding the number of "fancy" diamonds where its price per carat is at least $10000 and its average price:

```{r}
summarise(diamonds, num_fancy_diamonds = sum(price/carat > 10000), mean_diamond_price = mean(price))

```
The number of fancy diamonds that exceeds over $10,000$ is around 617, and the average price of a diamond is at $3932.8$

Determining if there are any diamonds of at least one carat that cost less than $1000:

```{r}
summarise(diamonds, num_cheap_diamonds = sum(carat >= 1 & price < 1000), mean_diamond_price = mean(price))
```
From the table, it seems that there are no diamonds that are less than $1000.

## 4 Multi-step transformations
### 4.1 The pipe (%>%)

```{r pipe}
diamonds %>% # pipe in the data
filter(cut == "Premium") %>% # restrict to premium cut diamonds
mutate(price_per_carat = price/carat) %>% # add price_per_carat variable
arrange(desc(price_per_carat)) # sort based on price_per_carat

```

Plotting a histogram to further visualize the data:
```{r}
diamonds %>% # pipe in the data
filter(cut == "Premium") %>% # restrict to premium cut diamonds
mutate(price_per_carat = price/carat) %>% # add price_per_carat variable
ggplot() + # start a ggplot
geom_histogram(aes(x = price_per_carat))
```

Calculating the mean price for diamonds of volume at least one carat:
```{r}
diamonds %>%
  filter(carat >= 1) %>%
  summarise(mean_price_one_carat = mean(price))
```
The average price of one carat is at $8142.115.

### 4.2 group_by()

Finding the maximum price of diamonds for each cut:

```{r}
diamonds %>% 
group_by(cut) %>% 
summarise(max_price = max(price))

```

Finding the maximum price for each group by cut and clarity:

```{r}
diamonds %>% # pipe in the data
group_by(cut, clarity) %>% # group by both cut and clarity
summarise(max_price = max(price))
```

Ungrouping the data, we get:

```{r}
diamonds %>% # pipe in the data
group_by(cut, clarity) %>% # group by both cut and clarity
summarise(max_price = max(price)) %>% # find the max price for each group
ungroup()

```

```{r}
count(diamonds, cut)
```

Reproducing the output of count(diamonds, cut) via group_by and summarise()
```{r}
diamonds %>%
  group_by(cut) %>%
  summarise(count = n())

```
Majority of the diamonds are categorized as **Ideal** and **Premium**, and the least number of diamonds are in the **Poor** category.

## Exercises

**What is the minimum diamond price in the dataset? See if you can find the answer in two different ways.**

Using summarise() and min() functions:

```{r}
diamonds %>%
  summarise(min_price = min(price))

```

Using arrange() and head():
```{r}
diamonds %>%
  arrange(price) %>%
  head(1)
```

The minimum diamond price in the dataset is $326.

**How many diamonds have length at least one and a half times their width?**

```{r}
diamonds %>%
  filter(x >= 1.5 * y) %>%
  summarise(count = n())
```
There are 10 diamonds that have a length at least one and a half times their width. 

**Among diamonds with colors D, E, F, G, what is the median number of carats for diamonds of each cut?**
```{r}
diamonds %>%
  filter(color %in% c("D", "E", "F", "G")) %>%
  group_by(cut) %>%
  summarise(median_carat = median(carat))
```
The table above shows the median number of carats for diamonds of each cut with colors D, E, F, and G. **Fair cut** has a median of **0.91**, **Good cut** has 0.72, **Very Good** has **0.70**, **Premium** has **0.71**, and **Ideal** has **0.52**. It appears that fair has the high number of median, while ideal has the lowest.  
